{\bf {\Large Four decades of water quality change in the upper San Francisco Estuary}}

### 2017-10-03 sftrends_manu.Rnw
better intro maybe done

\section*{Abstract}
\noindent Long-term monitoring data from the Delta region of the \ac{sfe} were modeled with an estuarine adaptation of the \ac{wrtds} to describe historical trends and relationships between key species of dissolved inorganic nitrogen (ammonium, nitrate/nitrite, total).  Trend analysis with flow-normalized results demonstrated the potential to misinterpret changes using observed data that include flow effects, such that several trends with flow-normalized data had changes in magnitude and even reversal of trends relative to the observed.  We further described mechanisms of change with two case studies that 1) evaluate downstream changes in nitrogen following upgrades at a wastewater treatment plant, and 2) interactions between biological invaders, chlorophyll, and flow in Suisun Bay.  \ac{wrtds} results for ammonium trends showed a distinct signal as a result of upstream \ac{wwtp} upgrades, with specific reductions observed in the winter months during low-flow conditions. Results for Suisun Bay showed that \ac{chla} production in early years was directly stimulated by flow, whereas the relationship with flow in later years was indirect and confounded by grazing pressure. Although these trends and potential causes of change have been described in the literature, results from \ac{wrtds} provided an approach to test alternative hypotheses of spatiotemporal drivers of nutrient dynamics in the Delta.
\acresetall

@@

The Sacramento - San Joaquin River Delta (hereafter `Delta') is a mosaic of inflows upstream of the \ac{sfe} that receives and processes inputs from the Central Valley watershed \citep{Jassby00,Jassby02,Jassby08}. Sediment export downstream of the Delta and \ac{wwtp} inputs are primary sources of nutrients for the larger Bay.  Background nutrient concentrations in \ac{sfe} often exceed those associated with excessive primary production, although ecosystem responses symptomatic of eutrophication have historically been infrequent. Changes in response to stressors suggests that recent conditions in \ac{sfe} have not followed past trajectories. For example, recent increases in phytoplankton biomass, reductions in dissolved oxygen, and increasing abundance of species associated with harmful algal blooms motivated the proposal of \ac{chla} thresholds to assess and manage eutrophication in South Bay \citep{Lehman05,Cloern07,Sutula17}. These changes stem from multiple potential sources, although variation in freshwater inputs/withdrawals, establishment of invasive species, and effects of climate change have been implicated as key factors \citep{Cloern12b}.  Many of these changes are linked to inputs from the Delta region.  \citet{Glibert14} described recent phytoplankton blooms in Suisun Bay that were attributed to increased residence times and increased rates of nitrification that occurred during a drought period.  Changes in flow management practices compounded with climate variation have altered flushing rates and turbidity as key factors that moderate phytoplankton growth in the system \citep{Alpine92,Lehman00,Wright04,Canuel09}.

Formal methods for trend analysis methods could help describe recent changes in water quality in the Delta region. Direct evaluation of observed time series is often insufficient, given that a long-term change can be masked by variation at shorter time scales or the observed variation represents the combined effects of many variables \citep{Oneill89,Levin92}.  As a practical approach for water quality evaluation, trend analysis of ecosystem response indicators often focuses on tracking the change in concentrations or loads of nutrients over many years. Response indicators can vary naturally with changing flow conditions and may also reflect long-term effects of management or policy changes. For example, \ac{chla} concentration as a measure of phytoplankton response to nutrient inputs can follow seasonal patterns with cyclical variation in temperature and light changes throughout each year, whereas annual trends can follow long-term variation in nutrient inputs to the system \citep{Cloern96,Cloern10}. Similarly, nutrient trends that vary with hydrologic loading also vary as a function of utilization rates by primary producers or decomposition processes \citep{Sakamoto89,Schultz08,Harding16}.

The \ac{wrtds} approach was developed in this context and has been used to characterize decadal trends in running-water systems \citep{Hirsch10,Sprague11,Medalie12,Hirsch14,Pellerin14,Zhang16}.  The \ac{wrtds} method has been adapted for trend analysis in tidal waters, with a focus on \ac{chla} trends in Tampa Bay \citep{Beck15} and the Patuxent River Estuary \citep{Beck17}. The goal of this study was to provide a comprehensive description of nutrient trends in the northern \ac{sfe} and Delta region to inform understanding of ecosystem response dynamics and potential causes of water quality change. We applied an estuarine adaptation of \ac{wrtds} to describe nitrogen trends in  different spatial and temporal contexts. The specific objectives were to \begin{inparaenum}[1\upshape)]  
\item quantify and interpret trends over four decades at ten stations in the Delta, including annual, seasonal, and spatial changes in nitrogen analytes and response to flow variation, and
\item provide detailed descriptions of two case studies in the context of conceptual relationships modeled with \ac{wrtds}.
\end{inparaenum}
The second objective evaluated two specific water quality stations as additional case studies to demonstrate complexities with nutrient response to flow, effects of nutrient-related source controls on ambient conditions, and effects of biological invasion by benthic filter feeders on primary production.  Our general hypothesis was that the results were expected to support previous descriptions of trends in this well-studied system, but that new insight into spatial and temporal variation in response endpoints was expected, particuarly in flow-normalized model predictions.

@@

The role of nutrients in stimulating primary production in \ac{sfe} has been the focus of several recent investigations\ \citep{Dugdale07,Parker12,Glibert14}. % see glibert14 description

@@

Nutrient time series of monthly observations from 1976 to 2013 were obtained for ten sampling stations (\cref{fig:delt_map}, \url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}, \cite{IEP13}).  Stations were grouped by location in the study area for comparison: peripheral Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; interior Delta stations D19, D26, D28; and Suisun stations D4, D6, and D7.  These stations were chosen based on continuity of the water quality time series and significance of their geographic location for understanding regional trends.  Time series were complete for all stations except for an approximate ten year gap from 1996-2014 for D19.  Data were minimally processed, with the exception of averaging replicates that occurred on the same day.  The three nitrogen analytes that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two). Less than 3\% of all observations were left-censored, although variation was observed between analytes and location.  The ammonium time series had the most censored observations at sites \Sexpr{cvals$Site_Code[1]} (\Sexpr{cvals$cper[1]}\% of all observations), \Sexpr{cvals$Site_Code[2]} (\Sexpr{cvals$cper[2]}\%), \Sexpr{cvals$Site_Code[3]} (\Sexpr{cvals$cper[3]}\%), \Sexpr{cvals$Site_Code[4]} (\Sexpr{cvals$cper[4]}\%), and \Sexpr{cvals$Site_Code[5]} (\Sexpr{cvals$cper[5]}\%).

>>>>

\section*{Abstract}
\noindent Long-term monitoring data from the Delta region of the \ac{sfe} were modeled with an estuarine adaptation of the \ac{wrtds} to describe historical trends and relationships between key species of dissolved inorganic nitrogen (ammonium, nitrate/nitrite, total).  Trend analysis with flow-normalized results demonstrated the potential to misinterpret changes using observed data that included flow effects, such that several trends with flow-normalized data had changes in magnitude and even reversal of trends relative to the observed.  We further described mechanisms of change with two case studies that evaluated 1) downstream changes in nitrogen following upgrades at a wastewater treatment plant, and 2) interactions between biological invaders, chlorophyll, and flow in Suisun Bay.  \ac{wrtds} results for ammonium trends showed a distinct signal as a result of upstream \ac{wwtp} upgrades, with specific reductions observed in the winter months during low-flow conditions. Results for Suisun Bay showed that \ac{chla} production in early years was directly stimulated by flow, whereas the relationship with flow in later years was indirect and confounded by grazing pressure. Although these trends and potential causes of change have been described in the literature, results from \ac{wrtds} provided an approach to test alternative hypotheses of spatiotemporal drivers of nutrient dynamics in the Delta.
\acresetall

@@

The Sacramento - San Joaquin River Delta (hereafter `Delta') is a mosaic of inflows upstream of the \ac{sfe} that receives and processes inputs from the Central Valley watershed \citep{Jassby00,Jassby02,Jassby08}. Sediment transport and \ac{wwtp} inputs from the Delta are primary sources of nutrients for the larger Bay.  Although water quality conditions in \ac{sfe} symptomatic of eutrophication have historically been infrequent, changes in response to stressors suggests that recent conditions have not followed past trajectories. For example, recent increases in phytoplankton biomass, reductions in dissolved oxygen, and increasing abundance of species associated with harmful algal blooms motivated the proposal of \ac{chla} thresholds to assess and manage eutrophication in the lower bay \citep{Lehman05,Cloern07,Sutula17}. Although these changes are potentially linked to drivers at different spatial and temporal scales, inputs from the Delta remain a critical interest for understanding downstream effects.

Nutrient concentrations are generally non-limiting for phytoplankton growth in the upper \ac{sfe}.  In contrast, light conditions have long been considered the primary limiting factor preventing accumulation of phytoplankton biomass \citep{Cole84,Alpine88}, whereas grazing pressure from pelagic and benthic species can reduce phytoplankton during periods of growth \citep{Nichols85,Jassby08,Kimmerer14}.  Moreover, changes in flow management practices compounded with climate variation have altered flushing rates and turbidity as key factors that moderate phytoplankton growth in the system \citep{Alpine92,Lehman00,Wright04,Canuel09}.  \citet{Glibert14} described recent phytoplankton blooms in Suisun Bay that were attributed to increased residence times and increased rates of nitrification that occurred during a drought period.  Changes in phytoplankton biomass and the effects on pelagic organisms have also been of concern.  \citet{Jassby08} described decadal trends in phytoplankton biomass to understand mechanisms of decline for pelagic fish populations in the Delta.  Although phytoplankon concentrations have been relatively consistent in Suisan Bay, biomass in the upper Delta has been increasing.  Much of these trends were explained by invasion of benthic grazers in polyhaline areas and changes in the mean flow conditions observed in the Delta.

A comprehensive water quality monitoring program has been in place in the Delta for several decades \citep{IEP13}.  Although these data have been used extensively, a systematic assessment of trends covering the full spatial and temporal coverage of the monitoring dataset has not been made.  This information could inform both past and current research by providing a cohesive description of system-wide trends in nutrients.  However, formal methods for trend analysis are required given that long-term changes can be masked by variation at shorter time scales or the observed variation represents the combined effects of many variables \citep{Oneill89,Levin92}.  As a practical approach for water quality evaluation, trend analysis of ecosystem response indicators often focuses on tracking the change in concentrations or loads of nutrients over many years. Response indicators can vary naturally with changing flow conditions and may also reflect long-term effects of management or policy changes. For example, \ac{chla} concentration as a measure of phytoplankton response to nutrient inputs can follow seasonal patterns with cyclical variation in temperature and light changes throughout each year, whereas annual trends can follow long-term variation in nutrient inputs to the system \citep{Cloern96,Cloern10}. Similarly, nutrient trends that vary with hydrologic loading also vary as a function of utilization rates by primary producers or decomposition processes \citep{Sakamoto89,Schultz08,Harding16}.

The \ac{wrtds} approach was developed in this context and has been used to characterize decadal trends in running-water systems \citep{Hirsch10,Sprague11,Medalie12,Hirsch14,Pellerin14,Zhang16}.  The \ac{wrtds} method has been adapted for trend analysis in tidal waters, with a focus on \ac{chla} trends in Tampa Bay \citep{Beck15} and the Patuxent River Estuary \citep{Beck17}. The goal of this study was to provide a comprehensive description of nutrient trends in the northern \ac{sfe} and Delta region to inform understanding of ecosystem response dynamics and potential causes of water quality change. The specific objectives were to \begin{inparaenum}[1\upshape)]  
\item quantify and interpret trends over four decades at ten stations in the Delta, including annual, seasonal, and spatial changes in nitrogen analytes and response to flow variation, and
\item provide detailed descriptions of two case studies in the context of conceptual relationships modeled with \ac{wrtds}.
\end{inparaenum}
The second objective evaluated two specific water quality stations to demonstrate complexities with nutrient response to flow, effects of nutrient-related source controls on ambient conditions, and effects of biological invasion by benthic filter feeders on primary production.  Our general hypothesis was that the results were expected to support previous descriptions of trends in this well-studied system, but that new insight into spatial and temporal variation in response endpoints was expected, particuarly in flow-normalized model predictions.

@@

@@

Nutrient time series of monthly observations from 1976 to 2013 were obtained for ten sampling stations (\cref{fig:delt_map}, \url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}, \cite{IEP13}).  Stations were grouped by location in the study area for comparison: peripheral Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; interior Delta stations D19, D26, D28; and Suisun stations D4, D6, and D7.  These stations include all but two of the long-term monitoring stations from the Delta and cover all of the major inflow and outflows \citep{Jabusch09}.  Time series were complete for all stations except for an approximate ten year gap from 1996-2014 for D19.  Data were minimally processed, with the exception of averaging replicates that occurred on the same day.  The three nitrogen analytes that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two). Less than 3\% of all observations were left-censored, although variation was observed between analytes and location.  The ammonium time series had the most censored observations at sites \Sexpr{cvals$Site_Code[1]} (\Sexpr{cvals$cper[1]}\% of all observations), \Sexpr{cvals$Site_Code[2]} (\Sexpr{cvals$cper[2]}\%), \Sexpr{cvals$Site_Code[3]} (\Sexpr{cvals$cper[3]}\%), \Sexpr{cvals$Site_Code[4]} (\Sexpr{cvals$cper[4]}\%), and \Sexpr{cvals$Site_Code[5]} (\Sexpr{cvals$cper[5]}\%).

### 2017-10-03 sftrends_manu.tex
better intro maybe done

\begin{singlespace}
\title{{\bf {\Large Four decades of water quality change in the upper San Francisco Estuary}}}
\author{
  {\bf {\normalsize Marcus W. Beck$^{1*}$, Thomas W. Jabusch$^2$, Philip R. Trowbridge$^2$, David B. Senn$^2$}}
  \\\\{\textit {\normalsize $^*$Corresponding author: \href{mailto:marcusb@sccwrp.org}{marcusb@sccwrp.org}}}
  \\\\{\textit {\normalsize $^1$USEPA National Health and Environmental Effects Research Laboratory}}
  \\{\textit {\normalsize Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}}
  \\\\{\textit {\normalsize Current address: Southern California Coastal Water Research Project}}
  \\{\textit {\normalsize 3535 Harbor Blvd, Suite 110, Costa Mesa, CA 92626}}
  \\\\{\textit {\normalsize $^2$San Francisco Estuary Institute}}
	\\{\textit {\normalsize 4911 Central Ave, Richmond, CA 94804}}
  \vspace{1in}
  \\ Version Date:   Thu Sep 28 11:58:28 2017 -0700
	}
\date{}
\maketitle
\end{singlespace}
\clearpage

\section*{Abstract}
\noindent Long-term monitoring data from the Delta region of the \ac{sfe} were modeled with an estuarine adaptation of the \ac{wrtds} to describe historical trends and relationships between key species of dissolved inorganic nitrogen (ammonium, nitrate/nitrite, total).  Trend analysis with flow-normalized results demonstrated the potential to misinterpret changes using observed data that include flow effects, such that several trends with flow-normalized data had changes in magnitude and even reversal of trends relative to the observed.  We further described mechanisms of change with two case studies that 1) evaluate downstream changes in nitrogen following upgrades at a wastewater treatment plant, and 2) interactions between biological invaders, chlorophyll, and flow in Suisun Bay.  \ac{wrtds} results for ammonium trends showed a distinct signal as a result of upstream \ac{wwtp} upgrades, with specific reductions observed in the winter months during low-flow conditions. Results for Suisun Bay showed that \ac{chla} production in early years was directly stimulated by flow, whereas the relationship with flow in later years was indirect and confounded by grazing pressure. Although these trends and potential causes of change have been described in the literature, results from \ac{wrtds} provided an approach to test alternative hypotheses of spatiotemporal drivers of nutrient dynamics in the Delta.
\acresetall

@@

The role of nutrients in stimulating primary production in \ac{sfe} has been the focus of several recent investigations\ \citep{Dugdale07,Parker12,Glibert14}.

@@

Nutrient time series of monthly observations from 1976 to 2013 were obtained for ten sampling stations (\cref{fig:delt_map}, \url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}, \cite{IEP13}).  Stations were grouped by location in the study area for comparison: peripheral Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; interior Delta stations D19, D26, D28; and Suisun stations D4, D6, and D7.  These stations were chosen based on continuity of the water quality time series and significance of their geographic location for understanding regional trends.  Time series were complete for all stations except for an approximate ten year gap from 1996-2014 for D19.  Data were minimally processed, with the exception of averaging replicates that occurred on the same day.  The three nitrogen analytes that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two). Less than 3\% of all observations were left-censored, although variation was observed between analytes and location.  The ammonium time series had the most censored observations at sites C10 (25.4\% of all observations), MD10 (18.1\%), D28 (17.8\%), D19 (12\%), and D7 (7.9\%).

>>>>

\begin{singlespace}
\title{{\bf {\Large Four decades of water quality change in the upper San Francisco Estuary}}}
\author{
  {\bf {\normalsize Marcus W. Beck$^{1*}$, Thomas W. Jabusch$^2$, Philip R. Trowbridge$^2$, David B. Senn$^2$}}
  \\\\{\textit {\normalsize $^*$Corresponding author: \href{mailto:marcusb@sccwrp.org}{marcusb@sccwrp.org}}}
  \\\\{\textit {\normalsize $^1$USEPA National Health and Environmental Effects Research Laboratory}}
  \\{\textit {\normalsize Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}}
  \\\\{\textit {\normalsize Current address: Southern California Coastal Water Research Project}}
  \\{\textit {\normalsize 3535 Harbor Blvd, Suite 110, Costa Mesa, CA 92626}}
  \\\\{\textit {\normalsize $^2$San Francisco Estuary Institute}}
	\\{\textit {\normalsize 4911 Central Ave, Richmond, CA 94804}}
  \vspace{1in}
  \\ Version Date:   Fri Sep 29 12:02:06 2017 -0700
	}
\date{}
\maketitle
\end{singlespace}
\clearpage

\section*{Abstract}
\noindent Long-term monitoring data from the Delta region of the \ac{sfe} were modeled with an estuarine adaptation of the \ac{wrtds} to describe historical trends and relationships between key species of dissolved inorganic nitrogen (ammonium, nitrate/nitrite, total).  Trend analysis with flow-normalized results demonstrated the potential to misinterpret changes using observed data that included flow effects, such that several trends with flow-normalized data had changes in magnitude and even reversal of trends relative to the observed.  We further described mechanisms of change with two case studies that evaluated 1) downstream changes in nitrogen following upgrades at a wastewater treatment plant, and 2) interactions between biological invaders, chlorophyll, and flow in Suisun Bay.  \ac{wrtds} results for ammonium trends showed a distinct signal as a result of upstream \ac{wwtp} upgrades, with specific reductions observed in the winter months during low-flow conditions. Results for Suisun Bay showed that \ac{chla} production in early years was directly stimulated by flow, whereas the relationship with flow in later years was indirect and confounded by grazing pressure. Although these trends and potential causes of change have been described in the literature, results from \ac{wrtds} provided an approach to test alternative hypotheses of spatiotemporal drivers of nutrient dynamics in the Delta.
\acresetall

@@

@@

Nutrient time series of monthly observations from 1976 to 2013 were obtained for ten sampling stations (\cref{fig:delt_map}, \url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}, \cite{IEP13}).  Stations were grouped by location in the study area for comparison: peripheral Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; interior Delta stations D19, D26, D28; and Suisun stations D4, D6, and D7.  These stations include all but two of the long-term monitoring stations from the Delta and cover all of the major inflow and outflows \citep{Jabusch09}.  Time series were complete for all stations except for an approximate ten year gap from 1996-2014 for D19.  Data were minimally processed, with the exception of averaging replicates that occurred on the same day.  The three nitrogen analytes that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two). Less than 3\% of all observations were left-censored, although variation was observed between analytes and location.  The ammonium time series had the most censored observations at sites C10 (25.4\% of all observations), MD10 (18.1\%), D28 (17.8\%), D19 (12\%), and D7 (7.9\%).
